A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-22 06-appendix.tex
Conserta a definição de \pot' no apêndice

  \begin{case}
    $p=\ceu{p'\AtLoop{p''}}$.  There are three subcases.
    \begin{subcase}
      $p'=\ceu{\Nop}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-nop}.
      Hence~$p_1=p_2=\ceu{\Loop{p''}}$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $p'=\ceu{\Break}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-break}.
      Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $p'\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-adv}.
      Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
      \begin{align*}
        \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
        \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
      \end{align*}
      for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By the induction
      hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
      Hence~$p_1=\ceu{p_1'\AtLoop{p''}}=\ceu{p_2'\AtLoop{p''}}=p_2$
      and~$e_1=e_1'=e_2'=e_2$.
    \end{subcase}
  \end{case}

@@

  \begin{case}
    $p=\ceu{p'\AtAnd{p''}}$.  There are two subcases.
    \begin{subcase}
      $\isblocked(p',n)$~is false.  There are three subcases.
      \begin{subsubcase}
        $p'=\ceu{\Nop}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop1}.
        Hence~$p_1=p_2=p''$ and~$e_1=e_2\nil$.
      \end{subsubcase}
      \begin{subsubcase}\label{lem.det-nst.and-brk1}
        $p'=\ceu{\Break}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk1}.
        Hence~$p_1=p_2=\ceu{\clear(p'');\Break}$ and~$e_1=e_2\nil$.
      \end{subsubcase}
      \begin{subsubcase}\label{lem.det-nst.and-adv1}
        $p'\ne\ceu{\Nop},\ceu{\Break}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv1}.
        Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
        \begin{align*}
          \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
          \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
        \end{align*}
        for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By the induction
        hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
        Hence~$p_1=\ceu{{p_1'}\And{p''}}=\ceu{{p_2'}\And{p''}}=p_2$
        and~$e_1=e_1'=e_2'=e_2$.
      \end{subsubcase}
    \end{subcase}
    \begin{subcase}
      $\isblocked(p',n)$~is true.  There are three subcases.
      \begin{subsubcase}
        $p''=\ceu{\Nop}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop2}.
        Hence~$p_1=p_2=p'$ and~$e_1=e_2\nil$.
      \end{subsubcase}
      \begin{subsubcase}\label{lem.det-nst.and-brk2}
        $p''=\ceu{\Break}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk2}.
        Hence~$p_1=p_2=\ceu{\clear(p');\Break}$ and~$e_1=e_2=\nil$.
      \end{subsubcase}
      \begin{subsubcase}\label{lem.det-nst.and-adv2}
        $p''\ne\ceu{\Nop},\ceu{\Break}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv2}.  Thus there
        are derivations~$\pi_1''$ and~$\pi_2''$ such that
        \begin{align*}
          \pi_1''&\Vdash\<p'',n,\nil>\nst\<p_1'',n,e_1''>\\
          \pi_2''&\Vdash\<p'',n,\nil>\nst\<p_2'',n,e_2''>
        \end{align*}
        for some~$p_1'',p_2''\in\P$ and~$e_1'',e_2''\in\E$.  By the
        induction hypothesis, $p_1''=p_2''$ and~$e_1''=e_2''$.
        Hence~$p_1=\ceu{{p'}\And{p_1''}}=\ceu{{p'}\And{p_2''}}=p_2$
        and~$e_1=e_1''=e_2''=e_2$.
      \end{subsubcase}
    \end{subcase}
  \end{case}

@@

\begin{definition}\label{def.pot}
  The potency of a program~$p$ in reaction to event~$e$,
  denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ expressions
  that can be reached during a reaction of~$p$ to~$e$.  Formally,
  $\pot(p,e)$ is defined as follows:
  \[
    \pot(p,e)=\pot'(\bcast(p,e))\,,
  \]
  where
  \begin{enumerate}[(a)]
  \item$\pot'(p)=1$, if~$p=\ceu{\EmitInt(e)}$;
    %%
  \item$\pot'(p)=\max\{\pot'(p_1),\pot'(p_2)\}$,\\
    if~$p=\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}}$;
    %%
  \item$\pot'(p)=\pot'(p_1)$, if~$p=\ceu{\Loop{p_1}}$
    or~$p=\ceu{\Every{e}\,{p_1}}$;
    %%
  \item$\pot'(p)=\pot'(p_1)+\pot'(p_2)$, if~$p=p_1;\,p_2$
    or~$p=\ceu{p_1\AtLoop{p_2}}$
    with~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
    or~$p=\ceu{{p_1}\And{p_2}}$ or~$p=\ceu{{p_1}\Or{p_2}}$
    or~$p=\ceu{{p_1}\AtAnd{p_2}}$ or~$p=\ceu{{p_1}\AtOr{p_2}}$;
    %%
  \item$\pot'(p)=0$, otherwise.
  \end{enumerate}
  Note that~$\pot'$ only considers the~$\ceu{\EmitInt}$ expressions that
  occur before the first~$\ceu{\AwaitExt}$ in a given execution path.  This
  is because such~$\ceu{\AwaitExt}$ act as barriers that cannot be consumed
  in the reaction to the original event~$e$.
\end{definition}

\begin{definition}\label{def.rank}
  The \emph{rank} of a description~$\delta=\<p,n,e>$,
  denoted~$\rank(\delta)$, is a pair of nonnegative integers~$\<i,j>$ such
  that~$i=\pot(p,e)$, and $j=n$~if~$e=\nil$ or~$j=n+1$ otherwise.
\end{definition}

@@

\begin{lemma}\strut
  \begin{enumerate}[(a)]
  \item If~$\delta\outpush\delta'$ then~$\rank(\delta)=\rank(\delta')$.
  \item If~$\delta\outpop\delta'$ then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate}
\end{lemma}
\begin{proof}\strut
  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$,
  $\rank(\delta)=\<i,j>$, and~$\rank(\delta')=\<i',j'>$.
  \begin{enumerate}[(a)]
  \item Suppose
    \[
      \<p,n,e>\outpush\<p',n',e'>\,.
    \]
    By~\R{push}, $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
    Definition~\ref{def.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
    as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.

    It remains to be shown that~$i=i'$:
    \begin{align*}
      i&=\pot(p,e)
         \tag*{by Definition~\ref{def.rank}}\\
       &=\pot'(\bcast(p,e))
         \tag*{by Definition~\ref{def.pot}}\\
       &=\pot'(p')
         \tag*{since~$p'=\bcast(p,e)$}\\
       &=\pot'(\bcast(p',\nil))
         \tag*{by definition of~$\bcast$}\\
       &=\pot'(\bcast(p',e'))
         \tag*{since~$e'=\nil$}\\
       &=\pot(p',e')
         \tag*{by Definition~\ref{def.pot}}\\
       &=i'
         \tag*{by Definition~\ref{def.rank}}
    \end{align*}

  \item Suppose
    \[
      \<p,n,e>\outpop\<p',n',e'>\,.
    \]
    By~\R{pop}, $p=p'$, $n>0$, $n'=n-1$, and~$e=e'=\nil$.
    %%
    By Definition~\ref{def.pot}, $\pot(\bcast(p,e))=\pot(\bcast(p',e')$;
    hence~$i=i'$.  And by Definition~\ref{def.rank}, $j=n$, as~$e=\nil$,
    and~$j'=n-1$, as~$e'=\nil$ and~$n'=n-1$; hence~$j>j'$.
    Therefore,~${\<i,j>}>{\<i',j'>}$.\qedhere
  \end{enumerate}
\end{proof}

\begin{lemma}\label{lem.rank-nst}
If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{lemma}
\begin{proof}
  By induction on the structure of~$\nst$ derivations.
  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$, $\rank(\delta)=\<i,j>$,
  and~$\rank(\delta')=\<i',j'>$.  Suppose there is a derivation~$\pi$
  such that
  \[
    \pi\Vdash\<p,n,e>\nst\<p',n',e'>\,.
  \]
  By definition of~$\nst$, we have that~$e=\nil$ and $n=n'$.  Depending on
  the structure of program~$p$, there are~11 possibilities.  In each one of
  them, we show that~$\rank(\delta)\ge\rank(\delta')$.

  \begin{case}
    $p=\ceu{\Mem(id)}$.
    %%
    Then~$\pi$ is an instance of~\R{mem}.  Hence~$p'=\ceu{\Nop}$
    and~$e'=\nil$.  Thus~$\rank(\delta)=\rank(\delta')=\<0,n>$.
  \end{case}

  \begin{case}
    $p=\ceu{\EmitInt(e_1)}$.
    %%
    Then~$\pi$ is an instance of~\R{emit-int}.  Hence~$p'=\ceu{\Nop}$
    and~$e'=e_1$.
    Thus~${\rank(\delta)={\<1,n>}}>{\<0,n+1>=\rank(\delta')}$.
  \end{case}

  \begin{case}
    $p=\ceu{\CanRun(n)}$.
    %%
    Then~$\pi$ is an instance of~\R{can-run}.  Hence~$p'=\ceu{\Nop}$
    and~$e'=\nil$.  Thus~$\rank(\delta)=\rank(\delta')=\<0,n>$.
  \end{case}

  \begin{case}
    $p=\ceu{\IfElse{p}{p_1}{p_2}}$.
    %%
    There are two subcases.
    \begin{subcase}
      \label{lem.rank-nst.if-true}
      $\eval(\ceu{\Mem(\Id)})$~is true.
      %%
      Then~$\pi$ is an instance of~\R{if-true}.  Hence~$p'=\ceu{p_1}$
      and~$e'=\nil$.  Thus~$\rank(\delta)=\<i,n>$
      and~$\rank(\delta')=\<j,n>$, where~$i=\max\{pot'(p_1),pot'(p_2)\}$
      and~$j=\pot'(p_1)$.  Therefore~$i>j$.
    \end{subcase}
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$~is false.
      %%
      Similar to Case~\ref{lem.rank-nst.if-true}.
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{p_1;\,p_2}$.
    %%
    There are three subcases.
    \begin{subcase}
      $p_1=\ceu{\Nop}$.
      %%
      Then~$\pi$ is an instance of~\R{seq-nop}.  TODO.
    \end{subcase}
    \begin{subcase}
      $p_1=\ceu{\Break}$.
      %%
      Then~$\pi$ is an instance of~\R{seq-brk}.  TODO.
    \end{subcase}
    \begin{subcase}
      \label{lem.rank-nst.seq-adv}
      $p_1\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      Then~$\pi$ is an instance of~\R{seq-adv}.
      Thus there is a derivation~$\pi'$ such that
      \[
        \pi'\Vdash\<p_1,n,\nil>\nst\<p_1',n,e_1'>\,,
      \]
      for some~$p_1'\in\P$ and~$e_1'\in\E$.  Thus~$p'=p_1';p_2$
      and~$e'=e_1'$.  By the induction hypothesis,
      \begin{equation}
        \label{lem.rank-nst.seq-adv.eq1}
        \rank(\<p_1,n,\nil>)\ge\rank(\<p_1',n,e_1'>)\,.
      \end{equation}
      There are two subcases.
      \begin{subsubcase}
        $e'=\nil$.
        %%
        Then
        \begin{align*}
          \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\,,\enspace\text{and}\\
          \rank(\delta')&=\<\pot'(p_1')+\pot'(p_2),n>\,.
        \end{align*}
        By~\eqref{lem.rank-nst.seq-adv.eq1} and from the fact
        that~$e_1'=\nil$, $\pot'(p_1)\ge\pot'(p_1')$.  Therefore,
        $\rank(\delta)\ge\rank(\delta')$.
      \end{subsubcase}
      \begin{subsubcase}
        $e'\ne\nil$.
        %%
        Then~$\pi'$ is an instance of~\R{emit-int}, as no other rule affects
        the third component of the output description.
        \[
          \rank(\delta)=\<i,n>
          >\<i-1,n+1>
          =\rank(\delta')\,.
        \]
      \end{subsubcase}
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{\Loop{p_1}}$.
  \end{case}

  TODO\dots
\end{proof}

>>>>

  \begin{case}
    $p=\ceu{p'\AtLoop{p''}}$.  There are three subcases.
    \begin{subcase}
      $p'=\ceu{\Nop}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-nop}.
      Hence~$p_1=p_2=\ceu{\Loop{p''}}$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $p'=\ceu{\Break}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-brk}.
      Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $p'\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-adv}.
      Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
      \begin{align*}
        \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
        \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
      \end{align*}
      for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By the induction
      hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
      Hence~$p_1=\ceu{p_1'\AtLoop{p''}}=\ceu{p_2'\AtLoop{p''}}=p_2$
      and~$e_1=e_1'=e_2'=e_2$.
    \end{subcase}
  \end{case}

@@

  \begin{case}
    $p=\ceu{p'\AtAnd{p''}}$.  There are two subcases.
    \begin{subcase}
      $\isblocked(p',n)$~is false.  There are three subcases.
      \begin{subsubcase}
        $p'=\ceu{\Nop}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop1}.
        Hence~$p_1=p_2=p''$ and~$e_1=e_2\nil$.
      \end{subsubcase}
      \begin{subsubcase}
        \label{lem.det-nst.and-brk1}
        $p'=\ceu{\Break}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk1}.
        Hence~$p_1=p_2=\ceu{\clear(p'');\Break}$ and~$e_1=e_2\nil$.
      \end{subsubcase}
      \begin{subsubcase}\label{lem.det-nst.and-adv1}
        $p'\ne\ceu{\Nop},\ceu{\Break}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv1}.
        Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
        \begin{align*}
          \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
          \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
        \end{align*}
        for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By the induction
        hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
        Hence~$p_1=\ceu{{p_1'}\And{p''}}=\ceu{{p_2'}\And{p''}}=p_2$
        and~$e_1=e_1'=e_2'=e_2$.
      \end{subsubcase}
    \end{subcase}
    \begin{subcase}
      $\isblocked(p',n)$~is true.  There are three subcases.
      \begin{subsubcase}
        $p''=\ceu{\Nop}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop2}.
        Hence~$p_1=p_2=p'$ and~$e_1=e_2\nil$.
      \end{subsubcase}
      \begin{subsubcase}\label{lem.det-nst.and-brk2}
        $p''=\ceu{\Break}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk2}.
        Hence~$p_1=p_2=\ceu{\clear(p');\Break}$ and~$e_1=e_2=\nil$.
      \end{subsubcase}
      \begin{subsubcase}\label{lem.det-nst.and-adv2}
        $p''\ne\ceu{\Nop},\ceu{\Break}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv2}.  Thus there
        are derivations~$\pi_1''$ and~$\pi_2''$ such that
        \begin{align*}
          \pi_1''&\Vdash\<p'',n,\nil>\nst\<p_1'',n,e_1''>\\
          \pi_2''&\Vdash\<p'',n,\nil>\nst\<p_2'',n,e_2''>
        \end{align*}
        for some~$p_1'',p_2''\in\P$ and~$e_1'',e_2''\in\E$.  By the
        induction hypothesis, $p_1''=p_2''$ and~$e_1''=e_2''$.
        Hence~$p_1=\ceu{{p'}\And{p_1''}}=\ceu{{p'}\And{p_2''}}=p_2$
        and~$e_1=e_1''=e_2''=e_2$.
      \end{subsubcase}
    \end{subcase}
  \end{case}

@@

\begin{definition}
  \label{def.pot}
  %%
  The potency of a program~$p$ in reaction to event~$e$,
  denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ expressions
  that can be reached during a reaction of~$p$ to~$e$.

  More formally,
  \[
    \pot(p,e)=\pot'(\bcast(p,e))\,,
  \]
  where~$\pot'$ is an auxiliary function that counts the number of
  reachable~$\ceu{\EmitInt}$ in the program resulting from the broadcast of
  event~$e$ to~$p$.

  The auxiliary function~$\pot'$ is defined by the following clauses:
  \begin{enumerate}[(a)]
    \item$\pot'(\ceu{\EmitInt}(e))=1$.
    \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})
      =\max\{\pot'(p_1),\pot'(p_2)\}$.
    \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$.
    \item If~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
      \begin{align*}
        \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\
        %%
        \pot'(\ceu{p_1\AtLoop{p_2}})&=
        \begin{cases}
          \pot'(p_1)              &\text{if~(\dag)}\\
          \pot'(p_1)+\pot'(p_2)   &\text{otherwise}\\
        \end{cases}\\
        %%
        \pot'(\ceu{{p_1}\And{p_2}})&=\pot'(p_1)+\pot'(p_2)\\
        %%
        \pot'(\ceu{{p_1}\AtAnd{p_2}})&=\pot'(p_1)+\pot'(p_2)\\
        %%
        \pot'(\ceu{{p_1}\Or{p_2}})&=\pot'(p_1)+\pot'(p_2)\\
        %%
        \pot'(\ceu{{p_1}\AtOr{p_2}})&=\pot'(p_1)+\pot'(p_2)\,,
      \end{align*}
      where~(\dag) stands for: ``A~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$
      occurs in all execution paths of~$p_1$.''
      \item Otherwise, if none of~(a)--(d) applies, $\pot(\ast)=0$.
    \end{enumerate}

  TODO: Define~$\pot'(\ceu{\Every{e}\,{p_1}})$ above.
\end{definition}

\begin{definition}\label{def.rank}
  The \emph{rank} of a description~$\delta=\<p,n,e>$,
  denoted~$\rank(\delta)$, is a pair of nonnegative integers~$\<i,j>$ such
  that
  \begin{alignat*}{2}
    i&=\pot(p,e) &\quad\text{and}\quad
    j&=
       \begin{cases}
         n  &\text{if~$e=\nil$}\\
         n+1&\text{otherwise\,.}
       \end{cases}
  \end{alignat*}
\end{definition}

@@

\begin{lemma}\strut
  \begin{enumerate}[(a)]
  \item If~$\delta\outpush\delta'$ then~$\rank(\delta)=\rank(\delta')$.
  \item If~$\delta\outpop\delta'$ then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate}
\end{lemma}
\begin{proof}\strut
  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$,
  $\rank(\delta)=\<i,j>$, and~$\rank(\delta')=\<i',j'>$.
  \begin{enumerate}[(a)]
  \item Suppose~$\<p,n,e>\outpush\<p',n',e'>$.  Then, by~\R{push},
    $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
    Definition~\ref{def.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
    as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.
    %%
    It remains to be shown that~$i=i'$:
    \begin{align*}
      i&=\pot(p,e)
         \tag*{by Definition~\ref{def.rank}}\\
       &=\pot'(\bcast(p,e))
         \tag*{by Definition~\ref{def.pot}}\\
       &=\pot'(p')
         \tag*{since~$p'=\bcast(p,e)$}\\
       &=\pot'(\bcast(p',\nil))
         \tag*{by definition of~$\bcast$}\\
       &=\pot'(\bcast(p',e'))
         \tag*{since~$e'=\nil$}\\
       &=\pot(p',e')
         \tag*{by Definition~\ref{def.pot}}\\
       &=i'
         \tag*{by Definition~\ref{def.rank}}
    \end{align*}
    Therefore, $\<i,j>=\<i',j'>$.

  \item Suppose~$\<p,n,e>\outpop\<p',n',e'>$.  Then, by~\R{pop}, $p=p'$,
    $n>0$, $n'=n-1$, and~$e=e'=\nil$.
    %%
    By Definition~\ref{def.pot}, $\pot(\bcast(p,e))=\pot(\bcast(p',e'))$;
    hence~$i=i'$.  And by Definition~\ref{def.rank}, $j=n$, as~$e=\nil$,
    and~$j'=n-1$, as~$e'=\nil$ and~$n'=n-1$; hence~$j>j'$.
    Therefore,~${\<i,j>}>{\<i',j'>}$.\qedhere
  \end{enumerate}
\end{proof}

\begin{lemma}
  \label{lem.rank-nst}
  %%
  If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{lemma}
\begin{proof}
  We proceed by induction on the structure of~$\nst$ derivations.
  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$, $\rank(\delta)=\<i,j>$,
  and~$\rank(\delta')=\<i',j'>$.  By the theorem hypothesis, there is a
  derivation~$\pi$ such that
  \[
    \pi\Vdash\<p,n,e>\nst\<p',n',e'>\,.
  \]
  By definition of~$\nst$, $e=\nil$ and $n=n'$.  Depending on the structure
  of program~$p$, there are~11 possibilities.  In each one of them, we show
  that~$\rank(\delta)\ge\rank(\delta')$.

  \begin{case}
    $p=\ceu{\Mem(id)}$.
    %%
    Then~$\pi$ is an instance of~\R{mem}.  Hence~$p'=\ceu{\Nop}$
    and~$e'=\nil$.  Thus
    \[
      \rank(\delta)=\rank(\delta')=\<0,n>\,.
    \]
  \end{case}

  \begin{case}
    $p=\ceu{\EmitInt(e_1)}$.
    %%
    Then~$\pi$ is an instance of~\R{emit-int}.  Hence~$p'=\ceu{\Nop}$
    and~$e'=e_1\ne\nil$.
    Thus
    \[
      {\rank(\delta)={\<1,n>}}>{\<0,n+1>=\rank(\delta')}\,.
    \]
  \end{case}

  \begin{case}
    $p=\ceu{\CanRun(n)}$.
    %%
    Then~$\pi$ is an instance of~\R{can-run}.  Hence~$p'=\ceu{\Nop}$
    and~$e'=\nil$.  Thus
    \[
      \rank(\delta)=\rank(\delta')=\<0,n>\,.
    \]
  \end{case}

  \begin{case}
    $p=\ceu{\IfElse{p}{p_1}{p_2}}$.
    %%
    There are two subcases.
    \begin{subcase}
      \label{lem.rank-nst.if-true}
      $\eval(\ceu{\Mem(\Id)})$~is true.
      %%
      Then~$\pi$ is an instance of~\R{if-true}.  Hence~$p'=\ceu{p_1}$
      and~$e'=\nil$.  Thus
      \begin{align*}
        \rank(\delta)&=\<\max\{pot'(p_1),pot'(p_2)\},n>\\
                     &\ge\<\pot'(p_1),n>=\rank(\delta')\,.
      \end{align*}
    \end{subcase}
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$~is false.
      %%
      Similar to Case~\ref{lem.rank-nst.if-true}.
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{p_1;\,p_2}$.
    %%
    There are three subcases.
    \begin{subcase}
      \label{lem.rank-nst.seq-nop}
      $p_1=\ceu{\Nop}$.
      %%
      Then~$\pi$ is an instance of~\R{seq-nop}.
      Hence~$p'=p_2$ and~$e'=\nil$.  Thus
      \begin{align*}
        \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\\
                     &\ge\<pot'(p_2),n>=\rank(\delta')\,.
      \end{align*}
    \end{subcase}
    \begin{subcase}
      \label{lem.rank-nst.seq-brk}
      $p_1=\ceu{\Break}$.
      %%
      Then~$\pi$ is an instance of~\R{seq-brk}.
      Hence~$p'=p_1$ and~$e'=\nil$.  Thus
      \[
        \rank(\delta)=\<0,n>=\rank(\delta')\,.
      \]
    \end{subcase}
    \begin{subcase}
      \label{lem.rank-nst.seq-adv}
      $p_1\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      Then~$\pi$ is an instance of~\R{seq-adv}.  Hence there is a
      derivation~$\pi'$ such that
      \[
        \pi'\Vdash\<p_1,n,\nil>\nst\<p_1',n,e_1'>\,,
      \]
      for some~$p_1'$ and~$e_1'$.  Thus~$p'=p_1';p_2$ and~$e'=e_1'$.  By the
      induction hypothesis,
      \begin{equation}
        \label{lem.rank-nst.seq-adv.eq1}
        \rank(\<p_1,n,\nil>)\ge\rank(\<p_1',n,e_1'>)\,.
      \end{equation}
      There are two subcases.
      \begin{subsubcase}
        $e'=\nil$.
        %%
        Then
        \begin{align*}
          \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\enspace\text{and}\\
          \rank(\delta')&=\<\pot'(p_1')+\pot'(p_2),n>\,.
        \end{align*}
        By~\eqref{lem.rank-nst.seq-adv.eq1}, $\pot'(p_1)\ge\pot'(p_1')$.
        Thus
        \[
          \rank(\delta)\ge\rank(\delta')\,.
        \]
      \end{subsubcase}
      \begin{subsubcase}
        $e'\ne\nil$.
        %%
        Then~$\pi'$ contain one application of~\R{emit-int}, which consumes
        one~$\ceu{\EmitInt(e')}$ expression from~$p_1$ and implies
        $\pot'(p_1)>\pot'(p_1')$.  Thus
        \begin{align*}
          \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\\
                       &>\<\pot'(p_1')+\pot'(p_2),n+1>=\rank(\delta')\,.
        \end{align*}
      \end{subsubcase}
    \end{subcase}
  \end{case}

  \begin{case}
    \label{lem.rank-nst.loop-expd}
    $p=\ceu{\Loop{p_1}}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-expd}.
    Hence~$p'=\ceu{p_1\AtLoop{p_1}}$ and~$e'=\nil$.
    %%
    TODO: Use condition~(\dag) in definition of~$\pot'$.
  \end{case}

  \begin{case}
    $p=\ceu{{p_1}\AtLoop{p_2}}$.
    %%
    There are three cases.
    \begin{subcase}
      $p_1=\ceu{\Nop}$.
      %%
      Similar to Case~\ref{lem.rank-nst.seq-nop}.
    \end{subcase}
    \begin{subcase}
      $p_1=\ceu{\Break}$.
      %%
      Similar to Case~\ref{lem.rank-nst.seq-brk}.
    \end{subcase}
    \begin{subcase}
      \label{lem.rank-nst.loop-adv}
      $p_1\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      Similar to Case~\ref{lem.rank-nst.seq-adv}.
    \end{subcase}
  \end{case}

  \begin{case}
    \label{lem.rank-nst.and.expd}
    $p=\ceu{{p_1}\And{p_2}}$.
    %%
    Then~$\pi$ is an instance of~\R{and-expd}.
    Hence~$p'=\ceu{{p_1}\AtAnd(\CanRun(n);\,p_2)}$ and~$e'=\nil$.
  \end{case}

  \begin{case}
    $p=\ceu{{p_1}\AtAnd{p_2}}$.
  \end{case}

  \begin{case}
    $p=\ceu{{p_1}\Or{p_2}}$.
  \end{case}

  \begin{case}
    $p=\ceu{{p_1}\AtOr{p_2}}$.
  \end{case}
\end{proof}
